 Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation

### 2021-06-15 analysis/paper/shceof.Rmd
Varios camios

The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.  \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993, \textcolor{red}{¿Alguna referencia más nueva?}].
Thus, the Real cEOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.

@@

The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equal.
\textcolor{red}{No me quedo claro ese "equal", ¿igual de probables/posibles/factibles?}.
To visualize that, Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.

(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º. \textcolor{red}{Rugs...}

@@

pattern <- eof_filtered$right %>% 
  .[era5[lev == 200], on = "time"] %>% 
  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
  .[, estimate := mean(estimate), by = .(lat, term)] %>%
  dcast(lon + lat ~ term, value.var = "estimate")

set.seed(42)
fake_sam <- eof_filtered$right %>% 
  copy() %>% 
  .[, eof := sample(eof, replace = TRUE)]

cor_with_eof2 <- fake_sam[cEOF2, on = "time"] %>% 
  sep_ReIm() %>% 
  .[, correlate(eof, hgt), by = part]

fake_sam <- fake_sam %>% 
  setnames("eof", "value") %>% 
  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]

@@

var <- fake_eof$sdev %>%
  copy() %>% 
    .[, setNames(paste0(EOF, " (", scales::percent(r2), ")"),
                 EOF)]
var <- c(var, setNames(c("SAM", "Imaginary", "Real"), c("SAM", "Imaginary", "Real")))

@@

  # geom_contour_tanaka has problem with the perfectly symemtric SAM
  # adding just a tiny bit of noise solves it. 
  .[EOF == "SAM", eof := eof + rnorm(.N, sd = 1e-5)] %>%

  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = eof, fill = ..level..), global.breaks = TRUE, 
                    breaks = AnchorBreaks(0, 0.5, exclude = 0)) +

@@

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
To this synthetic dataset (\textcolor{red}{No me suena mucho la palabra "synthetic", ¿se usa en este contexto?}) we then apply traditional EOF.
By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure \@ref(fig:fake-eof) row a).
While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so it needs two EOFs to be represented correctly.
Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

\textcolor{red}{Los titulos de la figura 11 superior se prestan a la confusión, ya que no es exactamente el SAM y no se entiende qué sería "Imaginary" y "Real"}

The three EOFs of the synthetic dataset are shown in Figure \@ref(fig:fake-eof) row b.
The synthetic SAM can be seen in the first EOF while the cEOF2 is explained by the second and third.
However, even though the synthetic SAM is perfectly zonally symmetric by construction, the leading EOF is not.
The EOF decomposition mixes this mode with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern (\textcolor{red}{REF}).
This happens because, although in this synthetic dataset the zonally symmetrical SAM and the cEOF2 are uncorrelated by construction, the sample correlation is not null \\textcolor{red}{No me quedo 100% clara esta parte, ¿podrías expresarla un poco distinto?}
.
In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part.
Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].

This suggests that the zonal asymmetries of the SON SAM could very well be mostly a statistical artefact.
Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.
## Tropical sources

@@

```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200) +
  scale_fill_divergent_discretised("SST",
                                   limits = c(-1, 1),
                                   guide = guide_colorsteps_bottom(20, order = 2)) 
```

The connections between cEOFs and Tropical sources is also assesed.
Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2. The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2). This pattern is almost canonically positive ENSO (\textcolor{red}{REF ENSO}) and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2). This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].

@@

```{r enso-phase, fig.cap = " ONI plotted against cEOF2 phase for th 1979 -- 2019 period. Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively. Black line is the fit ONI ~ sin(phase) computed by OLS weighted by the magnitude of cEOF2.", fig.width = width_column, fig.height=3}
gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[cEOF == "cEOF2"] %>% 
  # .[, hgt := Anomaly(hgt)] %>% 
  .[, arg := Arg(hgt)] %>%
  .[, mag := Mod(hgt)] %>%
  .[enso, on = "time"] %>%
  na.omit()

@@

@@

  scale_color_brewer(NULL, palette = "Dark2",
                     labels = c("TRUE"  = "|EOF2| > 50%",
                                "FALSE" = "|EOF2| < 50%")) +
  scale_shape_manual(NULL, values = c("TRUE" = 20,
                                      "FALSE" = 18),
                     labels = c("TRUE"  = "|EOF2| > 50%",
                                "FALSE" = "|EOF2| < 50%")) +
  scale_y_continuous("ONI") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi, pi, by = 45*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  coord_cartesian(xlim = c(-pi, pi)) +
  labs(subtitle = equation) +
  panel_background
```

@@

@@

(ref:pp-america-cap) Regression of SON mean precipitation anomalies in South America (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase. For the 1979 -- 2018 \textcolor{red}{¿Cambiaste el año o es un typo?} period.
Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.

@@

(ref:pp-oceania-cap) As Figure \@ref(fig:pp-america) but for Oceania.

(ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitation anomalies averaged between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b).

model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, data = gdata)
fstatistic <- summary(model)$fstatistic
p_value <- pf(fstatistic["value"], fstatistic["numdf"], fstatistic["dendf"], 
        lower.tail = FALSE)

equation <- latex2exp::TeX(paste0("$\\mathrm{Precipitation} = ", 
        signif(coef(model)[1], 2), "\\sin{(\\mathrm{phase})} + ", 
        signif(coef(model)[2], 2), "\\cos{(\\mathrm{phase})}$"), output = "text")

  scale_color_brewer(NULL, palette = "Dark2",
                     labels = c("TRUE"  = "|EOF2| > 50%",
                                "FALSE" = "|EOF2| < 50%")) +
  scale_shape_manual(NULL, values = c("TRUE" = 20,
                                      "FALSE" = 18),
                     labels = c("TRUE"  = "|EOF2| > 50%",
                                "FALSE" = "|EOF2| < 50%")) +
  scale_y_continuous("Precipitation anomaly") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi, pi, by = 30*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  coord_cartesian(xlim = c(-pi, pi)) +
  labs(subtitle = equation) +
  panel_background
```

To better understand the relationship between precipitation anomalies over Western Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight.
The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
To perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, it would be recommended to align one of the axis (either the Real or the Imaginary) to the phase that maximises the described relationship.

\textcolor{red}{¿Había alguna razón para poner SA antes que Oceanía? Sino invertiría el orden, porque creo que es más obvio lo que decis en Sudamérica sobre el sinusoidal fit, si mostras antes de lo de Oceanía.}

@@

To the extreme that the leading EOF of geopotential height with the cEOF2 removes results in an annular mode that has none of the typical zonal asymmetries of the observed SAM.\textcolor{red}{No me pareció del todo claro lo que estás queriendo expresar acá}

>>>>

The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.  \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@grytsai2011].
Thus, the Real cEOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.

@@

The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
To visualize that, Figure \@ref(fig:phase-histogram) shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
For instance, years with cEOF2 phase within 45º of 0º are nearest the "positive Real" phase.
About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
It could be seen that Imaginary phase is the most common phase and It is also the direction that has the maximum relationship with ENSO.
Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.

(ref:phase-histogram-cap) Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º. The small vertical lines near the horizontal axis mark the observations.

@@

pattern <- sams[term == "full" & lev == 200] %>% 
  .[era5[lev == 200], on = c("lev", "time")] %>% 
  .[, FitLm(hgt, estimate), by = .(lon, lat, lev)] %>% 
  rm_intercept() %>% 
  .[, estimate := mean(estimate), by = .(lat)] %>% 
  .[, term := NULL] %>% 
  setnames("estimate", "eof")

set.seed(42)

fake_sam <-  sams[term == "full" & lev == 200] %>% 
  copy() %>% 
  .[, value := sample(estimate, replace = TRUE)] %>% 
  .[, estimate := NULL] %>% 
  .[, term := NULL]

# complement <- function(x, y) {
#   
#   e <- residuals(lm(x ~ y))
#   
#   e*sd(x) + mean(x)
#   
# }
# 
# fake_sam <-  sams[term == "full" & lev == 200] %>% 
#   copy() %>% 
#   .[, value := complement(estimate, y)] %>% 
#   .[, estimate := NULL] %>% 
#   .[, term := NULL]

cor_with_eof2 <- fake_sam[cEOF2, on = "time"] %>% 
  sep_ReIm() %>% 
  .[, correlate(value, hgt), by = part]

fake_sam <- fake_sam %>% 
  .[pattern, on = "lev", allow.cartesian = TRUE] %>% 
  .[, sam := value*eof]

@@

var <- fake_eof$sdev %>%
  copy() %>% 
    .[, setNames(paste0(EOF, " (", scales::percent(r2), ")"),
                 EOF)]
var <- c(var, setNames(c("Annular Pattern", "Imaginary cEOF2", "Real cEOF2"), c("SAM", "Imaginary", "Real")))

@@

  # geom_contour_tanaka has problem with the perfectly symemtric SAM
  # adding just a tiny bit of noise solves it.

  .[, eof := scale(eof), by = EOF] %>% 
  .[EOF == "SAM", eof := eof + rnorm(.N, sd = 1e-5)] %>%
  .[EOF %in% c("EOF1", "EOF2"), eof := -eof] %>% 
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = eof, fill = ..level..), global.breaks = TRUE, 
                    breaks = AnchorBreaks(0, 0.5, exclude = 0)) +

@@

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated annular pattern, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric annylar pattern with random sign and value sampled with replacement from the real SAM index.
To this synthetic dataset we then apply traditional EOF.
By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure \@ref(fig:fake-eof) row a).
While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so traditional EOF analysis needs two EOFs to represented it correctly.
Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

The three EOFs of the synthetic dataset are shown in Figure \@ref(fig:fake-eof) row b.
The annular pattern is captured mainly by the first EOF while the cEOF2 is explained by the second and third.
However, even though the annular pattern used to construct these data is perfectly zonally symmetric, the leading EOF is not.
The EOF decomposition mixes the true annular pattern with parts of the the zonally asymmetric cEOF2 pattern, resulting in a pattern pretty similar to the real SAM pattern [@fogt2020].

\textcolor{red}{Todo esto está mal.}

Rotating makes it even worse.

The reason this happens is that the spatial correlation between the annular pattern and the wave patterns is not zero; i.e. they are not orthogonal. And since EOFs

In this particular example, it is `r cor_with_eof2[part == "Real"]$text` with the Real part and `r cor_with_eof2[part == "Imaginary"]$text` with the Imaginary part.
Of note, the three EOFs are well separated according to North's rule of thumb [@north1982].

This suggests that the zonal asymmetries of the SON SAM could very well be mostly a statistical artefact.
Since many surface impacts are mediated by the asymmetric component, as well as the relationship between SAM and ENSO [@campitelli2021], this potential issue could affect the interpretability of many results involving the SAM.
Studies on the relationship between the SAM and PSA-like patterns would be particularly difficult to interpret.

## Tropical sources

@@

```{r sst-psi-2,  fig.cap = "(ref:sst-psi-2-cap)", fig.width = 7, fig.height=4, dependson="plot_sst_psi", fig.env="figure*"}
plot_sst_psi(which_cEOF = "cEOF2", which_lev = 200) +
  scale_fill_divergent_discretised("SST",
                                   limits = c(-1, 1),
                                   guide = guide_colorstrip_bottom(20, order = 2)) 
```

The connections between cEOFs and Tropical sources is also assesed.
Figure \@ref(fig:sst-psi-2) shows the regression maps of Sea Surface Temperatures (SST) and Streamfunction at 200 hPa respectively upon cEOF2. The Imaginary cEOF2 is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2). This pattern is almost canonically positive ENSO [@bamston1997] and indeed, the correlation between the Imaginary cEOF2 and the Oceanic Niño Index [@bamston1997] is very high `r enso_cor[cEOF == "cEOF2" & part == "Imaginary"]$text`.
Besides, the Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2). This is consistent with the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites Rossby waves that propagate meridionally towards higher latitudes [@mo2000].

@@

```{r enso-phase, fig.cap = " ONI plotted against cEOF2 phase for th 1979 -- 2019 period. Years with magnitude of cEOF2 greater or smaller than the 50th percentile are shown as orange circles and green diamomnds respectively. Black line is the fit ONI ~ sin(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.", fig.width = width_column, fig.height=3}
gdata <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[cEOF == "cEOF2"] %>% 
  # .[, hgt := Anomaly(hgt)] %>% 
  .[, arg := Arg(hgt)] %>%
  .[, mag := Mod(hgt)] %>%
  .[enso, on = "time"] %>%
  na.omit()

@@

cor_enso_mag_quantile <- gdata[mag >= quantile(mag, 0.5)] %>% 
  .[oni < 1.5] %>% 
  .[, correlate(mag, abs(oni))]

@@

  scale_color_brewer(NULL, palette = "Dark2",
                     labels = c("TRUE"  = "|cEOF2| > 50%",
                                "FALSE" = "|cEOF2| < 50%")) +
  scale_shape_manual(NULL, values = c("TRUE" = 20,
                                      "FALSE" = 18),
                     labels = c("TRUE"  = "|cEOF2| > 50%",
                                "FALSE" = "|cEOF2| < 50%")) +
  scale_y_continuous("ONI") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi, pi, by = 45*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  coord_cartesian(xlim = c(-pi, pi)) +
  labs(subtitle = equation) +
  panel_background
```

@@

```{r}
# gdata %>% 
#   ggplot(aes(oni, mag)) +
#   geom_point(aes(color = mag >= quantile(mag, 0.5)))  +
#    scale_color_brewer(NULL, palette = "Dark2",
#                      labels = c("TRUE"  = "|cEOF2| > 50%",
#                                 "FALSE" = "|cEOF2| < 50%")) +
#   panel_background
```

@@

(ref:pp-oceania-cap) Regression of SON mean precipitation anomalies in Oceania (mm per day, shaded) and (column a) cEOF1 the (row 1) Imaginary and (column 1) Real phase. For the 1979 -- 2019.
Areas marked with dots have p-values smaller than 0.05 adjusted for False Detection Rate.

(ref:australia-pp-phase-cap) Same as Figure \@ref(fig:enso-phase) but for precipitation anomalies averaged between `r knitr::combine_words(LonLabel(box_australia$lon))` and `r knitr::combine_words(LatLabel(box_australia$lat))` (box shown in Figure \@ref(fig:pp-oceania) column b). 
Black line is the fit precipitation ~ sin(phase) + cos(phase) computed by weighted OLS using the magnitude of the cEOF2 as weights.

model <- lm(pp ~ I(sin(arg)) + I(cos(arg)) - 1, weights = mag, data = gdata)
fstatistic <- summary(model)$fstatistic
p_value <- pf(fstatistic["value"], fstatistic["numdf"], fstatistic["dendf"], 
        lower.tail = FALSE)

equation <- latex2exp::TeX(paste0("$\\mathrm{Precipitation} = ", 
        signif(coef(model)[1], 2), "\\frac{mm}{day} \\sin{(\\mathrm{phase})} + ", 
        signif(coef(model)[2], 2), "\\frac{mm}{day} \\cos{(\\mathrm{phase})}$"), output = "text")

  scale_color_brewer(NULL, palette = "Dark2",
                     labels = c("TRUE"  = "|EOF2| > 50%",
                                "FALSE" = "|EOF2| < 50%")) +
  scale_shape_manual(NULL, values = c("TRUE" = 20,
                                      "FALSE" = 18),
                     labels = c("TRUE"  = "|EOF2| > 50%",
                                "FALSE" = "|EOF2| < 50%")) +
  scale_y_continuous("Precipitation anomaly") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi, pi, by = 30*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  coord_cartesian(xlim = c(-pi, pi)) +
  labs(subtitle = equation) +
  theme(plot.subtitle = element_text(size = 8)) +
  panel_background
```

To better understand the relationship between precipitation anomalies over Eastern Australia and the cEOF2, Figure \@ref(fig:australia-pp-phase) plots precipitation anomalies averaged over Eastern Australia (box in Figure \@ref(fig:pp-oceania)) as a function of cEOF2 phase.
The black line shows a sinusoidal fit, whose coefficients are fitted by weighted least squares using the magnitude of the EFO2 as weight.
The $r^2$ corresponding to the fit is `r signif(summary(model)$r.squared, 2)` (p-value `r report_p(p_value)`).
There is a strong relationship between the phase of the cEOF2 and precipitation anomalies in Eastern Australia, but the Real and Imaginary phases we chose are not exactly aligned with the direction that maximises this relationship.
To perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, it would be recommended to align one of the axis (either the Real or the Imaginary) to the phase that maximises the described relationship.

(ref:pp-america-cap) As Figure \@ref(fig:pp-america) but for South America

@@

@@
